Carbonate clumped isotope constraints on Silurian ocean temperature and $\delta^{18}$O

### 2013-08-02 draft.tex
draft has been converted from word. All figures in good places. ready to send to Woody

\documentclass[5p, authoryear]{elsarticle}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{rotating}
\usepackage{multirow}
\usepackage{enumerate}
\usepackage{wasysym}
\usepackage{natbib}
\usepackage{lineno}
\usepackage{dblfloatfix}

@@

\newcommand{\deltao}{$\delta^{18}$}
\newcommand{\deltac}{$\delta^{13}$}
\newcommand{\degrees}{$^{\circ}$}

@@

\title{Carbonate clumped isotope constraints on Silurian ocean temperature and \deltao O}

@@

\begin{abstract}
Much of what we know about the history of Earth's climate derives from the geologic record of carbonates. The oxygen isotopic composition (\deltao O) of carbonates has been widely used as a proxy for past ocean temperatures, but application of this proxy to deep geologic time is complicated by diagenetic alteration and uncertainties in the \deltao O of seawater in the past. Carbonate clumped isotope thermometry provides an independent estimate of the temperature of the water from which a carbonate precipitated, allowing direct calculation of the \deltao O of the water as well. The clumped isotope composition of carbonates is also highly sensitive to diagenetic alteration, enabling identification of the most pristine samples from within a sample set. These highly pristine samples ultimately provide a more reliable record of paleoclimate.

We tested the ability of carbonate clumped isotopes to constrain paleoclimate by measuring the clumped isotope composition of apparently well-preserved calcitic fossils from several Silurian-age (ca. 427 Ma) stratigraphic sections on the island of Gotland, Sweden. Variability in clumped isotope temperatures suggests differential preservation, complicating a resolved temperature history through these sections. Micro-scale studies of calcite fabric and trace metal composition using electron backscatter diffraction and electron microprobe analysis point to a subset of relatively pristine samples, which indicate that Silurian tropical oceans were likely warm ($30\pm 2$\degrees C) and similar in oxygen isotopic composition to a modern ice-free world (\deltao O$_{\text{SMOW}}$ of $-1.7\pm 0.5\permil$). This result joins the growing body of evidence that suggests the $\delta^{18}$O of Earth's surface waters has remained broadly constant through time. 
\end{abstract}

@@

Despite the importance of paleoclimate and more than half a century of investigations on the topic, the temperature history of Earth's oceans remains poorly constrained. The most widely-used proxy for past ocean temperatures is the carbonate-water oxygen isotope exchange thermometer, also known as the carbonate \deltao O thermometer, which was developed by Harold Urey in the 1940s \citep{Urey1947}. This carbonate \deltao O proxy has two significant sources of complications. First, as with many paleoenvironmental proxies, it is often difficult to distinguish the effects of diagenetic alteration from the primary environmental signature. Second, the measured \deltao O of carbonates depends not only on the temperature of the seawater from which the carbonates precipitated, but also on the \deltao O of the seawater. The \deltao O of seawater can be affected by several factors, including evaporation, mixing with freshwater, and changes in the size of continental glaciers, all of which are difficult to quantify in the geologic record. The 
\deltao O of the ocean has also been hypothesized to change gradually but substantially through Earth's history \citep{Jaffres2007,Veizer1999}.

Gradual changes in the \deltao O of seawater on 100-million-year timescales seem plausible given the current knowledge of the controls on the \deltao O of seawater. The long-term average \deltao O of all the water on Earth's surface is almost entirely controlled by the balance between two exchange processes: exchange of oxygen between water and rocks during high-temperature alteration of the seafloor near mid-ocean ridges, and exchange of oxygen between water and rocks during low-temperature weathering of oceanic and continental crust. High-temperature exchange processes have small oxygen isotope fractionation factors and tend to increase the \deltao O of ocean water by equilibrating the water with the average mantle \deltao O of 5.7$\permil$ \citep{Gregory1991}. Low-temperature exchange processes have larger oxygen isotope fractionation factors and tend to decrease the \deltao O of seawater by preferentially moving $^{18}$O from water into rocks. The relative magnitudes of high- and low-temperature weathering are likely to have changed over the Phanerozoic due to uncoupled changes in the seafloor spreading rate \citep{Fornari1995,Tajika1993,Wallmann2001} and the magnitude of contintental weathering, which would lead to slow shifts in the \deltao O of the ocean. The best geologic evidence for gradual change in seawater \deltao O, however, is the subject of a long-standing controversy in geochemistry.

This controversy arises from the complications associated with the carbonate \deltao O proxy, which allow conflicting interpretations of the $\sim 8\permil$ increase in the average \deltao O of marine carbonates from the Cambrian to today. Several studies have proposed that early Paleozoic carbonates may have low \deltao O values because they are more likely to have undergone diagenetic alteration \citep{Clayton1959,Keith1964,Land1995}, which tends to decrease the \deltao O of carbonate \citep{Marshall1992}. In response to this concern, \cite{Veizer1999} measured the \deltao O of calcite fossils that were screened for diagenetic alteration using trace metal compositions and micro-scale studies of calcite fabrics. These screened fossils also record increasing \deltao O through the Phanerozoic, arguing against diagenetic alteration as a cause of the low \deltao O values in the early Paleozoic. An alternative hypothesis holds that oceans could have been much warmer in the early Paleozoic \citep{Karhu1986,Knauth1976}. Assuming that the \deltao O of the oceans was the same as today, the early Paleozoic oceans must have been $\sim 56$\degrees C to produce carbonates with a \deltao O$_{\text{PDB}}$ of $-8\permil$. These warm temperatures seem incompatible with evidence for a large glaciation in the Late Ordovician \citep{Veizer1986}, leading some studies to suggest a third hypothesis, that the \deltao O$_{\text{SMOW}}$ of seawater has increased from $-8\permil$ in the Cambrian to its present value of $\sim 0\permil$ \citep{Jaffres2007,Veizer1999}.

Carbonate clumped isotope thermometry provides a unique insight into the controversy described above because it resolves the two main complications of the carbonate \deltao O proxy. The clumped isotope composition of a carbonate records a temperature that is independent of the \deltao O of water and highly sensitive to diagenetic alteration. The extent to which $^{13}$C and $^{18}$O occur together, or ``clump'' within the same carbonate ion is a function of the temperature at which the carbonate precipitates. For a shelly marine organism that lives in shallow water, the degree of clumping in the shell calcite will reflect the sea surface temperature \citep{Ghosh2006}. Measurement of the \deltao O of the shell calcite in addition to the temperature allows calculation of the \deltao O of the seawater, using the experimentally-determined temperature dependence of the oxygen isotope fractionation between water and calcite \citep{Kim1997}. If the calcite shell is then buried and fossilized, it is often altered in the presence of high temperatures. This alteration can take the form of recrystallization or thermally-activated re-ordering \citep{Eiler2011}. In either case, the altered material will take on a new clumped-isotope temperature that can be much higher than the temperature of the pristine fossil calcite. If the altered material is mixed with pristine fossil material during sample processing, the clumped isotope temperature of the sample will be significantly higher than the temperature of the ocean water where the organism lived. We demonstrate here that carbonate clumped isotopes can be a more sensitive indicator of diagenetic alteration than the \deltao O or the trace metal content of a carbonate.

In addition to being highly sensitive to diagenetic alteration, carbonate clumped isotopes are also diagnostic of diagenetic regimes, ultimately allowing identification of the most pristine endmember of a sample set. Clumped isotopes can distinguish between diagenesis in the presence of low and high water-to-rock ratios (rock-buffered and water-buffered diagenesis, respectively) and diagenesis in the presence of meteoric water. As carbonates are buried deeper and the temperature increases, the equilibrium oxygen isotope fractionation between water and carbonate grows smaller, preferentially driving $^{18}$O into the water. During rock-buffered diagenesis, this causes the \deltao O of the water to increase, while the \deltao O of the altered carbonate is not affected. In a water-buffered diagenetic regime, the \deltao O of the water does not change, so increasing temperature instead causes the \deltao O of the altered carbonate to decrease. In contrast, diagenesis in the presence of meteoric water usually occurs when the sample is near the Earth's surface, so the clumped isotope temperature of the altered carbonate is cool, but the \deltao O of the carbonate and the water is diagnostically low. The ability of clumped isotopes to constrain the conditions to which fossils have been exposed after burial allows identification of the best-preserved samples from within a set of samples that traditionally would all be considered pristine \citep{Eiler2011}. We demonstrate here that this distinction has important implications for the estimation of the temperature and \deltao O of oceans in the past.

The present study investigates the implications of diagenetic alteration for the resolution of small temperature changes recorded in relatively well-preserved Paleozoic rocks. This study also provides a test of established proxies for diagenetic alteration as filters for clumped isotope data. Finally, our data add to the clumped isotope record of past ocean temperatures and \deltao O, including samples from a different continent and a different paleocean basin than previous clumped isotope studies.

@@

Samples were collected from a section exposed on the island of Gotland, Sweden, that spans the latest Llandovery through the Wenlock in the Early Silurian (Figure \ref{map}). Ages of the formations and correlations between outcrops are constrained by conodont and graptolite biostratigraphy \citep{Jeppsson2006}. The section records shelf, carbonate platform and backreef lagoon sediments deposited on the margin of the Baltic basin, which was within 30$^{\circ}$ of the equator during the Silurian \citep{Torsvik1992}.

\begin{figure*}[t]
\centering
\includegraphics[width=140mm]{Fig_01-01.png}
\caption{Geologic map of the northern half of the island of Gotland, Sweden, modified from \cite{Calner2004a}. Approximate sample locations are marked and labeled with location number, which corresponds to the sample locations in the supplementary table.}
\label{map}
\end{figure*}

The oldest exposed strata on Gotland are the Lower Visby Beds, which consist of marls interbedded with fine-grained limestones and were likely deposited below storm wave base (Figure \ref{column}). Fossils are present but not abundant \citep{Calner2004a,Samtleben1996}. The contact between Lower and Upper Visby beds is marked by a bed with a dense population of \textit{Phaulactis}, a species of solitary rugose coral \citep{Jeppsson1997,Jeppsson2006,Munnecke2003,Samtleben1996}. This contact is just above the Llandovery-Wenlock boundary, based on conodont stratigraphy \citep{Aldridge1993,Jeppsson1983,Mabillard1985}. During sampling, we defined this contact as a stratigraphic height of zero (Figure \ref{column}). In addition to the bed of abundant \textit{Phaulactis} corals, the contact between Lower and Upper Visby Beds is also marked by a facies change that represents a maximum flooding surface \citep{Calner2004b}. Carbonate content increases in the Upper Visby Beds, with a decreasing frequency of marl beds and the appearance of detritic limestones and reef mounds. Ripple marks indicate deposition above storm wave base. Fossils are abundant in the Upper Visby Beds \citep{Calner2004a,Samtleben1996}.

@@

Above the Upper Visby Beds, the H\"{o}gklint Formation consists of algal and crinoidal limestones with some reef mounds, deposited above storm wave base \citep{Riding1991,Samtleben1996,Watts2000}. The contact between the H\"{o}gklint Formation and the overlying Tofta Formation represents a hypothesized sequence boundary \citep{Calner2004b}. The Tofta Formation consists of oncoid-rich bedded limestone, and likely represents a shallow restricted setting, such as a backreef lagoon \citep{Riding1991,Samtleben1996}. The overlying Hangvar Formation consists of interbedded marls and limestones with some reef mounds, indicating deposition in a less restricted, deeper-water setting. The youngest strata sampled for this study were from the Slite Formation, which transitions from limestone-rich to marl-rich interbedded limestones and marls, likely representing deposition in deeper water. Fossils are present throughout the section \citep{Calner2004a}.

The Silurian section on Gotland is an attractive location to sample for a clumped isotope study for several reasons. First, this section has long been recognized for its excellent preservation. The strata on Gotland are very nearly flat-lying, with a dip of less than 1\degrees, and display only minimal evidence of tectonic disturbance \citep{Calner2004a,Jeppsson1983}. Conodont fossils from this section are pale yellow in color, giving them a conodont alteration index (CAI) of one, which corresponds to a maximum temperature of $\sim$100\degrees C \citep{Epstein1976,Jeppsson1983,Wenzel2000}.

The Silurian section on Gotland is also attractive due to the relative abundance of fossils, especially brachiopods, throughout the section. By sampling fossil material for clumped isotope measurements, we target calcite that was first precipitated directly from seawater. All brachiopod samples were taken from the secondary interior fibrous layer of the brachiopod shell, which is thought to precipitate in equilibrium with seawater and to be relatively resistant to diagenetic alteration \citep{Azmy1998,Samtleben2001}. The fibrous calcite fabric characteristic of the secondary layer has been well-characterized in modern and fossil brachiopods \citep{PerezHuerta2007,Samtleben2001,Schmahl2004}, providing an expectation for pristine samples in our dataset. The trace metal composition of brachiopods has also been extensively studied as an indicator of diagenetic alteration \citep{Azmy1998,Brand2003,Brand2012,Mii1994,Grossman1996,Shields2003}. The other fossils present in this section, including rugose corals, crinoids, bryozoans, and gastropods, are used less often for paleoenvironmental studies than brachiopods, but have been used by other clumped isotope studies of paleotemperature \citep{Came2007,Dennis2013,Finnegan2011}. Measurements of several fossil taxa allow comparison of the preservation quality of different calcite materials.

A third attractive feature of the Silurian section on Gotland is the coupled positive excursion in \deltac C and \deltao O across the boundary between the Lower and Upper Visby Beds (Figure \ref{column}). This excursion is called the Ireviken event, and is recognized globally \citep{Munnecke2003}. In Gotland, the 3$\permil$ excursion in \deltac C and 0.6$\permil$ excursion in \deltao O reach a maximum rate of change across the contact between the Lower and Upper Visby Beds \citep{Munnecke2003}. The excursions coincide with a shift to deposition in shallower water, but it is difficult to distinguish the relative contributions of progradation and eustatic sea level fall to this facies change \citep{Calner2004a}. The excursions are also broadly coincident with minor extinctions in several benthic and pelagic marine taxa \citep{Munnecke2003}.

The coincident isotopic excursions and stratigraphic change suggest that the Ireviken represents a climate phenomenon, the nature of which is still debated. The similarity between the isotopic signatures of the Ireviken event and known glaciations, such as the Late Ordovician glaciation, have led several studies to hypothesize a glacial cause for the Ireviken event \citep{Azmy1998,Kaljo2003,Brand2006,Calner2008}. This hypothesis is supported by the widespread presence of glacial tillites in Brazil, dated to the latest Llandovery or earliest Wenlock \citep{Grahn1992}. Many of these observations could also be explained by a shift from a humid period to an arid period in the tropics \citep{Bickert1997,Munnecke2003,Samtleben1996}. According to this model, during the humid period recorded by the Lower Visby Beds, increased precipitation would lower the \deltao O of surface waters and also stimulate the upwelling of deep water with low \deltac C onto the shelves. Increased precipitation would also cause intense weathering, which explains the dominance of marls over carbonate reefs during this time. Then during the arid period beginning in the Upper Visby Beds, evaporation would increase the \deltao O of surface waters, and downwelling of saline water would bring only surface water with high \deltac C onto the shelves. Low terrigenous influxes would allow the growth of carbonate reefs, observed in the H\"{o}gklint Formation.

In the present study, we test the ability of clumped isotopes to resolve the changes in ocean temperature and seawater \deltao O that contributed to the 0.6$\permil$ shift in the \deltao O of carbonates through the Ireviken event. From first principles, it will be difficult for the carbonate clumped isotope thermometer to resolve these changes, because a shift of 0.6$\permil$ in \deltao O corresponds to a change in temperature of approximately 2\degrees C, which is near the limit of resolution of carbonate clumped isotopes.

@@

Samples were collected from the island of Gotland, Sweden during the summer of 2011. Each sample was labeled with a location, as shown in Figure \ref{map}, and a stratigraphic height. For locations where the \textit{Phaulactis} layer is exposed, stratigraphic height was expressed relative to the boundary between the Lower and Upper Visby Beds (Figure \ref{column}). Other locations were correlated on the basis of prior sequence-, bio- and chemostratigraphic work \citep{Calner2004a}.

Most of the fossils were easily exposed by breaking the surrounding layers of poorly-lithified marl. For fossils in more strongly lithified beds, fossil material was accessed by cutting away the rock with a saw or microrotary drill. After fossils were separated from the rock, they were sonicated for two hours to remove residual sand, silt and mud from the surface of the fossils. Brachiopods, bryozoans, crinoids, gastropods and ostracods were sampled by flaking the surface of the skeleton or shell using a dental chisel. Every flake was visually inspected under a microscope at circa $25\times$ power. Flakes with visible pyrite grains or rough-textured recrystallized regions were rejected and not included in the material for analysis. For brachiopods, only the flakes preserving the characteristic fabric of the fibrous secondary layer were chosen for analysis (see Figure \ref{flakes}). Flakes were then sonicated briefly, rinsed, dried in a 50\degrees C oven, and powdered using a mortar and pestle. Only the apical ends of rugose corals were sampled, after removal of any sediment from the surface using an abrasive bit. The apical end was then cut off using a diamond microsaw, and powdered using a mortar and pestle. The apical ends of the corals were targeted because the coral septa are closely spaced together there, leaving less room for cement to precipitate between the septa. However, this sampling technique is still not as selective as the technique used for brachiopods, and may include more cement and sediment along with skeletal calcite. Finally, to assess diagenetic alteration, micrite and cement samples were drilled from cut surfaces and then powdered using a mortar and pestle.

>>>>

\documentclass[5p, authoryear]{elsarticle}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{rotating}
\usepackage{multirow}
\usepackage{enumerate}
\usepackage{wasysym}
\usepackage{natbib}
\usepackage{lineno}
\usepackage{dblfloatfix}
\usepackage[skip=2pt,font=footnotesize]{caption}
\usepackage{breakcites}

@@

@@

\title{Carbonate clumped isotope constraints on Silurian ocean temperature and $\delta^{18}$O}

@@

\begin{abstract}
Much of what we know about the history of Earth's climate derives from the geologic record of carbonates. The oxygen isotopic composition ($\delta^{18}$O) of carbonates has been widely used as a proxy for past ocean temperatures, but application of this proxy to deep geologic time is complicated by diagenetic alteration and uncertainties in the $\delta^{18}$O of seawater in the past. Carbonate clumped isotope thermometry provides an independent estimate of the temperature of the water from which a carbonate precipitated, allowing direct calculation of the $\delta^{18}$O of the water as well. The clumped isotope composition of carbonates is also highly sensitive to diagenetic alteration, enabling identification of the most pristine samples from within a sample set.

We evaluated preservation and measured the clumped isotope composition of a suite of calcitic fossils from Silurian (ca. 433 Ma) stratigraphic sections on the island of Gotland, Sweden. Substantial variability in clumped isotope temperatures suggests differential preservation, complicating the generation of a stratigraphically resolved temperature history through these sections. However, micro-scale studies of calcite fabric and trace metal composition using electron backscatter diffraction and electron microprobe analysis point to a subset of relatively pristine samples, which indicate that Silurian tropical oceans were likely warm (31$\pm$2$^{\circ}$C) and similar in oxygen isotopic composition to a modern ice-free world ($\delta^{18}$O$_{\text{VSMOW}}$ of -1.4$\pm$0.5$\permil$). This result joins the growing body of evidence that suggests the $\delta^{18}$O of Earth's surface waters has remained broadly constant through time. 
\end{abstract}

@@

Despite more than half a century of investigations on the topic, the long-term temperature history of Earth's oceans remains poorly constrained. The most widely used proxy for past ocean temperature is the carbonate-water oxygen isotope exchange thermometer, also known as the carbonate $\delta^{18}$O thermometer \citep{Urey1947}. This thermometer is based on the temperature dependence of the fractionation between the $\delta^{18}$O of carbonate, reported relative to the VPDB standard, and the $\delta^{18}$O of seawater, reported relative to the VSMOW standard. The carbonate $\delta^{18}$O proxy has two significant sources of complications. First, as with many paleoenvironmental proxies, it is often difficult to distinguish the effects of diagenetic alteration from the primary environmental signature. Second, the measured $\delta^{18}$O of carbonates depends not only on the temperature of the seawater from which the carbonates precipitated, but also on the $\delta^{18}$O of the seawater. The $\delta^{18}$O of seawater can be influenced by several processes, including local evaporation and mixing with meteoric water, as well as global changes in the size of continental glaciers, all of which are difficult to constrain in the geologic record.

The $\delta^{18}$O of the ocean has also been hypothesized to have changed gradually but substantially through Earth's history \citep{Jaffres2007, Veizer1999}. Gradual changes in the $\delta^{18}$O of seawater on 100-million-year timescales seem plausible given the current knowledge of the controls on the $\delta^{18}$O of seawater. The long-term average $\delta^{18}$O of all the water on Earth's surface is almost entirely controlled by the balance between two exchange processes: exchange of oxygen between water and rocks during high-temperature alteration of the seafloor near mid-ocean ridges, and exchange of oxygen between water and rocks during low-temperature weathering of oceanic and continental crust. High-temperature exchange processes have small oxygen isotope fractionation factors and tend to increase the $\delta^{18}$O of ocean water by equilibrating the water with the average mantle $\delta^{18}$O$_{\text{VSMOW}}$ of 5.7$\permil$ \citep{Gregory1991}. Low-temperature exchange processes have larger oxygen isotope fractionation factors and tend to decrease the $\delta^{18}$O of seawater by preferentially moving $^{18}$O from water into rocks. The relative magnitudes of high- and low-temperature weathering are likely to have changed over the Phanerozoic due to uncoupled changes in the seafloor spreading rate \citep{Fornari1995, Tajika1993, Wallmann2001} and the magnitude of continental weathering, which would lead to slow shifts in the $\delta^{18}$O of the ocean. The best geologic evidence for gradual change in seawater $\delta^{18}$O, however, is the subject of a long-standing controversy in geochemistry.

This controversy arises from the complications associated with the carbonate $\delta^{18}$O proxy, which allow conflicting interpretations of the $\sim$8$\permil$ increase in the average $\delta^{18}$O of marine carbonates from the Cambrian to today \citep{Jaffres2007, Veizer1999}. Several studies have proposed that early Paleozoic carbonates may have low $\delta^{18}$O values because they are more likely to have undergone diagenetic alteration \citep{Clayton1959, Keith1964, Land1995}, which tends to decrease the $\delta^{18}$O of carbonate \citep{Marshall1992}. In response to this concern, \cite{Veizer1999} measured the $\delta^{18}$O of calcite fossils that were screened for diagenetic alteration using trace metal compositions and micro-scale studies of calcite fabrics. These screened fossils also record increasing $\delta^{18}$O through the Phanerozoic, arguing against diagenetic alteration as a cause of the low $\delta^{18}$O values in the early Paleozoic. An alternative hypothesis holds that oceans could have been much warmer in the early Paleozoic \citep{Karhu1986, Knauth1976}. Assuming that the $\delta^{18}$O of the oceans was the same as today, the Cambrian oceans must have been $\sim$56$^{\circ}$C to produce carbonates with a $\delta^{18}$O$_{\text{VPDB}}$ of -8$\permil$. These warm temperatures seem incompatible with evidence for a large glaciation in the Late Ordovician \citep{Veizer1986}, leading some studies to suggest that the increase in the $\delta^{18}$O of marine carbonates reflects a concomitant increase in the $\delta^{18}$O$_{\text{VSMOW}}$ of seawater from -8$\permil$ in the Cambrian to its present value of $\sim$0$\permil$ \citep{Jaffres2007, Veizer1999}.

Carbonate clumped isotope thermometry provides a unique insight into this controversy because it provides data that can help to resolve the two main complications of the carbonate $\delta^{18}$O proxy: the clumped isotope composition of a carbonate records a temperature that is independent of the $\delta^{18}$O of water and highly sensitive to diagenetic alteration. The extent to which $^{13}$C and $^{18}$O occur together, or ``clump'' within the same carbonate ion is a function of the temperature at which the carbonate precipitates \citep{Eiler2011}, and isotopic clumping in shallow-water calcifiers has been shown to record ambient water temperatures \citep{Ghosh2006, Henkes2013, Saenger2012, Tripati2010}. Clumped isotope temperature estimates can in turn be used, with the $\delta^{18}$O of shell calcite and the experimentally determined temperature dependence of the oxygen isotope fractionation between water and calcite \citep{Kim1997}, to calculate the $\delta^{18}$O of the seawater.

Shells that have been deeply buried in lithified sediments have often experienced high-temperature alteration, which can change the $\delta^{18}$O of the carbonate and is frequently difficult to detect. Clumped isotopes are sensitive to this type of alteration because the altered material will take on a new clumped isotope temperature that can be much higher than the temperature of the pristine fossil calcite. If the altered material is incorporated within pristine fossil material selected for sample analysis (e.g. as cements lining fossil void space), the clumped isotope temperature of the sample will be significantly higher than the temperature of the ocean water where the organism lived. We demonstrate here that carbonate clumped isotopes can be a more sensitive indicator of diagenetic alteration than the $\delta^{18}$O or the trace metal content of a carbonate.

In addition to being highly sensitive to diagenetic alteration, carbonate clumped isotope arrays are also diagnostic of diagenetic regimes, ultimately allowing identification of the most pristine end-member of a sample set. Clumped isotopes can distinguish between diagenesis in the presence of low and high water-to-rock ratios (rock-buffered and water-buffered diagenesis, respectively) and diagenesis in the presence of meteoric water. As carbonates are buried deeper and the temperature increases, the equilibrium oxygen isotope fractionation between water and carbonate grows smaller, preferentially driving $^{18}$O into the water. During rock-buffered diagenesis, this causes the $\delta^{18}$O of the water to increase, while the $\delta^{18}$O of the altered carbonate is not affected. In a water-buffered diagenetic regime, increasing temperature instead causes the $\delta^{18}$O of the altered carbonate to decrease. In contrast, diagenesis in the presence of meteoric water usually occurs when the sample is near the Earth's surface, so the clumped isotope temperature of the altered carbonate is cool, but the $\delta^{18}$O of both the carbonate and the water are diagnostically low. The ability of clumped isotopes to constrain the conditions to which fossils have been exposed after burial allows identification of the best-preserved samples from within a set of samples that traditionally would all be considered pristine \citep{Eiler2011}. We demonstrate here that this distinction has important implications for the estimation of the temperature and $\delta^{18}$O of oceans in the past.

The present study investigates the implications of diagenetic alteration for the resolution of small temperature changes recorded in relatively well-preserved Paleozoic rocks. This study also provides a test of established proxies for diagenetic alteration as filters for clumped isotope data. Finally, our data add to the clumped isotope record of past ocean temperatures and $\delta^{18}$O, including samples from a different continent and a different paleocean basin than previous clumped isotope studies.

@@

Samples were collected from several sections exposed on the island of Gotland, Sweden that span the latest Llandovery through Wenlock epochs of the Silurian Period (Figure \ref{map}). Ages of the formations and correlations between outcrops are constrained by conodont and graptolite biostratigraphy \citep{Jeppsson2006}. The sections record shelf, carbonate platform and backreef lagoon sediments deposited on the margin of the Baltic basin, which was 19$^{\circ}$ south of the equator during the Silurian Period \citep{Torsvik1992}.

\begin{figure*}[t]
\centering
\includegraphics[width=140mm]{Fig_01-01.png}
\caption{Geologic map of the northern half of the island of Gotland, Sweden, modified from \cite{Calner2004a}. Approximate sample locations are marked and labeled with location number, which corresponds to the sample locations in Supplementary Table 1.}
\label{map}
\end{figure*}

The oldest exposed strata on Gotland are the Lower Visby Beds, which consist of marls interbedded with fine-grained limestones deposited below storm wave base (Figure \ref{column}). Fossils are abundant throughout the section \citep{Calner2004a, Samtleben1996}. The contact between Lower and Upper Visby beds is marked by a bed with abundant fossils of the solitary rugose coral genus \textit{Phaulactis} \citep{Jeppsson1997, Jeppsson2006, Munnecke2003, Samtleben1996}. This contact is just above the Llandovery-Wenlock boundary as defined by conodont biostratigraphy \citep{Aldridge1993, Jeppsson1983, Mabillard1985}. We defined the \textit{Phaulactis} bed as a datum and all stratigraphic positions are measured relative to it, except those in the Slite Formation, which are measured relative to an arbitrary bed in the Slite Formation. The contact between Lower and Upper Visby Beds is also characterized by a facies change that represents a maximum flooding surface \citep{Calner2004b}. Carbonate content increases in the Upper Visby Beds, with a decreasing frequency of marl beds and the appearance of bioclastic limestones and reef mounds. Ripple marks indicate deposition above storm wave base. Fossils are especially abundant in the Upper Visby Beds \citep{Calner2004a, Samtleben1996}.

@@

Above the Upper Visby Beds, the H\"{o}gklint Formation consists of algal and crinoidal limestones with some reef mounds, deposited above storm wave base \citep{Riding1991, Samtleben1996, Watts2000}. The contact between the H\"{o}gklint Formation and the overlying Tofta Formation represents a hypothesized sequence boundary \citep{Calner2004b}. The Tofta Formation is characterized by oncoid-rich bedded limestone, and likely represents a shallow restricted setting, such as a backreef lagoon \citep{Riding1991, Samtleben1996}. The overlying Hangvar Formation consists of interbedded marls and limestones with some reef mounds, indicating deposition in a less restricted, deeper-water setting. The youngest strata sampled for this study were from the Slite Formation, which transitions from limestone-rich to marl-rich interbedded limestones and marls, likely representing deposition in deeper water \citep{Calner2004a}.

The Silurian section on Gotland is an attractive location to sample for a clumped isotope study for several reasons. First, this section has long been recognized for its generally excellent preservation. The strata on Gotland are very nearly flat-lying, with a dip of less than 1$^{\circ}$, and display only minimal evidence of tectonic disturbance \citep{Calner2004a, Jeppsson1983}. Conodont fossils from this section yield a conodont alteration index (CAI) of one, which corresponds to a maximum temperature of $\sim$100$^{\circ}$C \citep{Epstein1976, Jeppsson1983, Wenzel2000}.

The Silurian section on Gotland is also attractive because of the abundance of fossils, especially brachiopods, throughout the section. As with previous paleoclimate studies we sampled fossil phases for clumped isotope measurements in order to target calcite that was first precipitated directly from seawater (as opposed to carbonate matrix, which may contain authigenic cements). Many studies of paleoenvironment have targeted the secondary interior fibrous layer of the brachiopod shell, which is thought to precipitate in equilibrium with seawater and to be relatively resistant to diagenetic alteration \citep{Azmy1998, Samtleben2001}. The fibrous calcite fabric characteristic of the secondary layer has been well characterized in modern and fossil brachiopods \citep{PerezHuerta2007, Samtleben2001, Schmahl2004}, providing a standard of comparison for the samples in our dataset. The trace metal composition of brachiopods has also been extensively studied as an indicator of diagenetic alteration \citep{Azmy1998, Brand2003, Brand2012, Mii1994, Grossman1996, Shields2003}. The other fossils present in this section, including rugose corals, crinoids, bryozoans, and gastropods, are used less often for paleoenvironmental studies than brachiopods, but have been used by other clumped isotope studies of paleotemperature \citep{Came2007, Dennis2013, Finnegan2011}. Measurements of several fossil taxa allow comparison of the relative preservation of different calcitic materials and can help to detect taxon-specific ``vital effects'' if they are present.

A third attractive feature of the Silurian section on Gotland is the coupled positive excursion in $\delta^{13}$C and $\delta^{18}$O that initiates at the boundary between the Lower and Upper Visby Beds (Figure \ref{column}). This excursion, termed the Ireviken event, is now recognized globally \citep{Munnecke2003}. In Gotland, a 3$\permil$ excursion in $\delta^{13}$C and 0.6$\permil$ excursion in $\delta^{18}$O attain a maximum rate of change across the contact between the Lower and Upper Visby Beds \citep{Munnecke2003}. The excursions coincide with a maximum flooding surface, but it is difficult to distinguish the relative contributions of progradation and eustatic sea level fall to this facies change \citep{Calner2004a}. The excursions are also broadly coincident with minor extinctions in several benthic and pelagic marine taxa \citep{Munnecke2003}.

The coincident isotopic excursions and stratigraphic change suggest that the Ireviken represents a climate phenomenon, the nature of which is still debated. The similarity between the isotopic signatures of the Ireviken event and known glaciations, such as the Late Ordovician glaciation, have led several studies to hypothesize a glacial cause for the Ireviken event \citep{Azmy1998, Kaljo2003, Brand2006, Calner2008}. This hypothesis is supported by the widespread presence of glacial tillites in Brazil that have been dated to the latest Llandovery or earliest Wenlock \citep{Grahn1992}. Many of these observations could also be explained by a shift from a humid period to an arid period in the tropics, which could accompany a glaciation or occur independently \citep{Bickert1997, Munnecke2003, Samtleben1996}. According to this model, during the humid period recorded by the Lower Visby Beds, increased precipitation would have lowered the $\delta^{18}$O of surface waters and also stimulated the upwelling of deep water with low $\delta^{13}$C onto the shelves. Increased precipitation would also have caused intense weathering, which could explain the dominance of marls over carbonate reefs during this time. During the succeeding arid period beginning in the Upper Visby Beds, evaporation would have increased the $\delta^{18}$O of surface waters, and downwelling of saline water would have brought surface water with high $\delta^{13}$C onto the shelves. Low terrigenous influxes would have allowed the growth of carbonate reefs as observed in the H\"{o}gklint Formation.

In the present study, we test the ability of clumped isotopes to resolve the changes in ocean temperature and seawater $\delta^{18}$O that contributed to the 0.6$\permil$ shift in the $\delta^{18}$O of carbonates through the Ireviken event. From first principles, it will be challenging but not impossible for the carbonate clumped isotope thermometer to resolve these changes, because a shift of 0.6$\permil$ in $\delta^{18}$O corresponds to a change in temperature of approximately 2$^{\circ}$C. Single measurements of carbonate clumped isotope composition can attain a resolution of $\pm$1$^{\circ}$C, while running multiple replicates of a sample can improve the resolution to $\pm$0.4$^{\circ}$C \citep{Thiagarajan2011}.

@@

Samples were collected from the island of Gotland, Sweden during the summer of 2011. Each sample was labeled with a location, as shown in Figure \ref{map}, and a stratigraphic height. For locations where the \textit{Phaulactis} layer is exposed, stratigraphic height was expressed relative to the boundary between the Lower and Upper Visby Beds (Figure \ref{column}). Other locations were correlated on the basis of prior sequence-, bio- and chemostratigraphic work \citep{Calner2004a}, as well as new bulk carbonate $\delta^{13}$C measurements (Fike et al. in prep).

Of the 22 brachiopods chosen for sampling, 20 belonged to the species group \textit{Atrypa reticularis}. We targeted this species because it is abundant in the sampled strata, the shell has a thick secondary fibrous layer, and it was used in several previous paleoenvironmental studies in Gotland \citep{Bickert1997, Munnecke2003, Samtleben2001}.

Most of the fossils were easily exposed by breaking the surrounding layers of poorly-lithified marl. For fossils in more strongly lithified beds, fossil material was accessed by cutting away the rock with a saw or microrotary drill. After fossils were separated from the rock, they were sonicated for two hours in a Cole Parmer 8891 ultrasonic cleaner to remove residual sand, silt and mud from the surface of the fossils. Brachiopods, bryozoans, crinoids, gastropods and ostracods were sampled by flaking the surface of the skeleton or shell using a dental chisel. Every flake was visually inspected under a microscope at circa $25\times$ power. Flakes with visible pyrite grains or rough-textured recrystallized regions were rejected and not included in the material for analysis. For brachiopods, only the flakes preserving the characteristic fabric of the fibrous secondary layer were chosen for analysis (see Figure \ref{flakes}). Flakes were then sonicated briefly, rinsed, dried in a 50$^{\circ}$C oven, and powdered using a mortar and pestle. Only the apical ends of rugose corals were sampled, after removal of any sediment from the surface using an abrasive bit. The apical end was then cut off using a diamond microsaw, and powdered using a mortar and pestle. The apical ends of the corals were targeted because the coral septa are closely spaced together there, leaving less room for cement to precipitate between the septa. However, this sampling technique is still not as selective as the technique used for brachiopods, and may include more cement and sediment along with skeletal calcite. Finally, to assess possible diagenetic end-members, micrite and cement samples were drilled from cut surfaces and then powdered using a mortar and pestle.
